### 2017-07-07 reasoning.tex
Edits.

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the transaction body ($c$).
The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_l$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step. If there is no such state $\stg_1$, then a transaction will never be able to take its next step after the interference, and hence such interference can be safely discarded.

@@

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of an
$\I$-constrained relation. We therefore introduce a side-condition on
$\I$ that restores the equality. The condition requires $\I$ to allow
an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
and $\stg''$, only if it also allows interference for every prefix of
$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
should not make the interference from $\stg$ to $\stg''$ valid.  This
stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid, then an interference has already modified a
record, and no further interference will restore the original record,
because the original record bears the id of a transaction that has
already committed. Thus, $\I_{ss}$ remains invalid.

@@

Salient rules of transaction-local RG judgment are shown in
Fig.~\ref{fig:rg-rules}. These rules (\rulelabel{RG-Update},
\rulelabel{RG-Select}, \rulelabel{RG-Delete}, and
\rulelabel{RG-Insert}) reflect the structure of the corresponding
reduction rule from Fig.~\ref{fig:txnimp}.  The rule
\rulelabel{RG-Foreach} defines the RG judgment for a \C{FOREACH} loop.
As is characteristic of loops, the reasoning is pivoted on a loop
invariant $\psi$ that needs to be stable w.r.t $\R$. $\psi$ must be
implied by $P$, the pre-condition of \C{FOREACH}, when no elements
have been iterated, i.e, when $y=\emptyset$. The body of the loop can
assume the loop invariant, and the fact that $z$ is an element from
the set $x$ being iterated, to prove its post-condition $Q_c$. The
operational semantics ensures that $z$ is added to $y$ at the end of
the iteration, hence $Q_c \conj z\in y$ is valid at the end of the
iteration, from which $\psi$ must follow. When the loop has finished
execution, $y$, the set of iterated items, is the entire set $x$. Thus
$\psi \conj y=x$ must imply the post-condition $Q$, which also needs
to be stable. The rules for conditionals, sequencing etc., are
standard, and hence elided.

@@

\begin{definition}[\bfseries Step-indexed reflexive transitive closure]
For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
\mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
the smallest relation satisfying the following
properties:
\begin{itemize}
\item $\forall (x:A).\, R^0 (x,x)$
\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
R^{n+1}(x,z)$
\end{itemize}
\end{definition}

\begin{definition}[\bfseries Interleaved step and multi-step relations]
Interleaved step relations interleave global and transaction-local
reductions with interference as captured by the corresponding rely
relations. They are defined thus:
\begin{smathpar}
\begin{array}{lclr}
(c,\stg) \rstepsto (c',\stg') & \defeq &  
  (c,\stg) \stepsto (c',\stg') \disj (c' = c \conj R(\stg, \stg'))&
  \texttt{global}\\
(c,\stl,\stg) \rstepsto (c',\stl',\stg') & \defeq & \stg \vdash 
  (c,\stl) \stepsto (c',\stl') \conj \stg'=\stg& \texttt{transaction-local}\\
  &   & \disj (c' = c \conj \stl'=\stl \conj \R(\stl, \stg, \stg'))
\end{array}
\end{smathpar}

\noindent An interleaved multi-step relation ($\stepssto{n}$) is the
step-indexed reflexive transitive closure of the interleaved step
relation.  
\end{definition}

\begin{definition}[\bfseries Semantics of RG judgments]
\label{def:rg-semantics}
The semantics of the global and transaction-local RG judgments are
defined thus:
\begin{smathpar}
\begin{array}{lcl}
\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~
  P(\stl,\stg) \conj (c,\stl,\stg) \rstepssto{n} (\cskip, \stl',\stg')
  \Rightarrow Q(\stl',\stg')\\
\rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
  \Rightarrow (\forall n,\stg'.\; (c,\stg) 
    \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
&&\hspace*{0.5in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
\end{array}
\end{smathpar}

@@

\begin{smathpar}
\begin{array}{lcl}
\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall n,\stg',c'',\stg''.
(c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto
  (c'',\stg'') \Rightarrow G(\stg',\stg'')\\
\end{array}
\end{smathpar}
\end{definition}

>>>>

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the body ($c$) of a transaction
$i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_l$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step. If there is no such state $\stg_1$, then a transaction will never be able to take its next step after the interference, and hence such interference can be safely discarded.

@@

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of an
$\I$-constrained relation. We therefore introduce a side-condition on
$\I$ that restores the equality. The condition requires $\I$ to allow
an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
and $\stg''$, only if it also allows interference for every prefix of
$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
should not make the interference from $\stg$ to $\stg''$ valid.  This
stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
modified a record, and no further interference will restore the
original record, because the original record bears the id of a
transaction that has already committed. Thus, $\I_{ss}$ remains
invalid.

@@

Salient rules of transaction-local RG judgment are shown in
Fig.~\ref{fig:rg-rules}. These rules (\rulelabel{RG-Update},
\rulelabel{RG-Select}, \rulelabel{RG-Delete}, and
\rulelabel{RG-Insert}) reflect the structure of the corresponding
reduction rule from Fig.~\ref{fig:txnimp}.  The rule
\rulelabel{RG-Foreach} defines the RG judgment for a \C{FOREACH} loop.
As is characteristic of loops, the reasoning is pivoted on a loop
invariant $\psi$ that needs to be stable w.r.t $\R$. $\psi$ must be
implied by $P$, the pre-condition of \C{FOREACH}, when no elements
have been iterated, i.e, when $y=\emptyset$. The body of the loop can
assume the loop invariant, and the fact that $z$ is an element from
the set $x$ being iterated, to prove its post-condition $Q_c$. The
operational semantics ensures that $z$ is added to $y$ at the end of
the iteration, hence $Q_c$ must imply $[y\cup\{z\}/y]\psi$. When the
loop has finished execution, $y$, the set of iterated items, is the
entire set $x$. Thus $[x/y]\psi$ is true at the end of the loop, from
which the post-condition $Q$ must follow. As with the other rules, $Q$
needs to be stable. The rules for conditionals, sequencing etc., are
standard, and hence elided.

@@

\begin{definition}[\bfseries Interleaved step and multi-step relations]
Interleaved step relations interleave global and transaction-local
reductions with interference as captured by the corresponding rely
relations. They are defined thus:
\begin{smathpar}
\begin{array}{lclr}
(c,\stg) \rstepsto (c',\stg') & \defeq &  
  (c,\stg) \stepsto (c',\stg') \disj (c' = c \conj R(\stg, \stg'))&
  \texttt{global}\\
(\tbox{c}_i,\stl,\stg) \rstepsto (\tbox{c'}_i,\stl',\stg') & \defeq & \stg \vdash 
  (\tbox{c}_i,\stl) \stepsto (\tbox{c'}_i,\stl') \conj \stg'=\stg& \texttt{transaction-local}\\
  &   & \disj (c' = c \conj \stl'=\stl \conj \R(\stl, \stg, \stg'))
\end{array}
\end{smathpar}

\noindent An interleaved multi-step relation ($\stepssto{n}$) is the
reflexive transitive closure of the interleaved step relation.  
\end{definition}

\begin{definition}[\bfseries Semantics of RG judgments]
\label{def:rg-semantics}
The semantics of the global and transaction-local RG judgments are
defined thus:
\begin{smathpar}
\begin{array}{lcl}
\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~
  P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
  (\tbox{\cskip}_i, \stl',\stg')
  \Rightarrow Q(\stl',\stg')\\
\rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
  \Rightarrow (\forall n,\stg'.\; (c,\stg) 
    \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
&&\hspace*{0.5in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
\end{array}
\end{smathpar}

@@

\begin{smathpar}
\begin{array}{lcl}
\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall \stg',c'',\stg''.
(c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto
  (c'',\stg'') \Rightarrow G(\stg',\stg'')\\
\end{array}
\end{smathpar}
\end{definition}
